Spatial Model

### 2019-12-10 rmds/linear_regression.Rmd
work from weekend at home

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

@@

# should find better way to read data from rmd folder
lm_dat <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/modeling-dat/lm_dat_logged.RDS')
```

@@

The simplest model we can fit is a linear regression. This model was inspired by Metro Transit's Transit Market Areas. Here, we'll let $Y_i$ equal the log of average number of weekday boardings per stop in census block group $i$ in 2017. We'll use four basic predictors, inspired by the original Transit Market Areas: population density, employment density, percent of households with no vehicle, and a walkability metric derived from walk isochrones. We also include number of stops made in block group $i$ as a control for level of service.

@@

```{r}
lm1 <- lm(log_daily_boards ~ pop_density + emp_density + walkability + perc_no_veh + log_daily_stops, lm_dat, na.action = na.exclude)

@@

I'm guessing population density is insignificant because it's so correlated with employment.

# Results
## Transit Market Areas

```{r}
lm_dat[, index := -3.79 + 0.06*pop_density + 0.22*emp_density + 0.21*walkability + 0.19*perc_no_veh + 1.18*log_daily_stops]

lm_dat[, predicted_boardings := exp(index)]

lm_dat_sf <- dplyr::left_join(bgs, lm_dat)
```

We can use the model above to define updated Transit Market Areas based on the number of boardings a block group can support, rather than an arbitary index value. The distribution of predicted boardings per stop is below:

```{r}
ggplot(lm_dat, aes(x=predicted_boardings)) + 
  geom_histogram() +
  theme_minimal() +
  labs(title = "Predicted Boardings")
```

```{r}
ggplot(lm_dat, aes(x=daily_boards, y = predicted_boardings)) +
  geom_point() +
  geom_smooth()

ggplot(lm_dat[daily_boards < 200], aes(x=daily_boards, y = predicted_boardings)) +
  geom_point() +
  geom_smooth()
```

There are several ways we can split this distribution to achieve 5 distinct market areas. However, a map of the continuous estimate is also informative:

```{r}
ggplot(lm_dat_sf %>% dplyr::filter(!is.na(predicted_boardings))) +
  geom_sf(aes(fill = predicted_boardings), lwd = 0) +
  scale_fill_viridis_c() +
  theme_minimal()
```

Here we can see the "bullseye" pattern we might expect

@@

>>>>

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

@@

# should find better way to read data from rmd folder
lm_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/lm_dat_scaled.RDS')
```

@@

The simplest model we can fit is a linear regression. This model was inspired by Metro Transit's Transit Market Areas. Here, we'll let $Y_i$ equal the log of average number of weekday boardings per stop in census block group $i$ in 2017. We'll use four basic predictors, inspired by the original Transit Market Areas: population density, employment density, percent of households with no vehicle, and a walkability metric derived from walk isochrones. We also include number of stops made in block group $i$ as a control for level of service. Each variable is scaled for comparison.

@@

```{r}
lm_dat <- lm_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(log_daily_stops)]
lm1 <- lm(scale(log_daily_boards) ~ pop_density + emp_density + perc_no_veh + walkability + scale(log_daily_stops), lm_dat, na.action = na.exclude)

@@

# Results

## Predictions

```{r}
lm_dat$predicted_log_scale_boards <- lm1$fitted.values
ggplot(lm_dat, aes(x=scale(log_daily_boards), y = predicted_log_scale_boards)) +
  geom_point() +
  labs(title = "Actual versus fitted") +
  theme_minimal() + 
  xlim(-3, 3) + ylim(-3, 3)
```

## Transit Market Areas
There are several ways we can split this distribution to achieve 5 distinct market areas. However, a map of the continuous estimate is also informative:

```{r}
lm_dat$resid <- residuals(lm1)
lm_dat_sf <- dplyr::left_join(bgs, lm_dat)

ggplot(lm_dat_sf %>% dplyr::filter(!is.na(predicted_log_scale_boards))) +
  geom_sf(aes(fill = predicted_log_scale_boards), lwd = 0) +
  scale_fill_viridis_c() +
  theme_minimal()
```

Here we can see the "bullseye" pattern we might expect.

@@

```{r}
ggplot(lm_dat_sf %>% dplyr::filter(!is.na(resid))) + 
  geom_sf(aes(fill = resid), lwd = 0) +
  scale_fill_viridis_c() + 
  theme_minimal() +
  labs(title = "residuals")
```

```{r}
library(spdep)
resid_dat <- lm_dat_sf %>%
  dplyr::filter(!is.na(resid))

nb <- poly2nb(resid_dat)
W <- nb2mat(nb, style = "B", zero.policy = TRUE)
listW <- nb2listw(nb, zero.policy = TRUE)

moran.mc(resid_dat$log_daily_boards, listW, nsim = 9999, zero.policy = TRUE)
```

```{r}
moran.mc(resid_dat$resid, listW, nsim = 9999, zero.policy = TRUE) # completely spatially random
```

Moran's I suggests some spatial clustering. The "strength" of clustering is less than we might expect because transit ridership occurs on linear paths rather than in "bullseye" clusters. That said, the observed rank in the Moran simulation above suggests that the remaining clustering is worth addressing with an inherently spatial model!

### 2019-12-10 rmds/poisson_regression.Rmd
work from weekend at home

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

# The Model

Let $Y_i$ equal the average number of boardings in block group $i$ in 2017, and let $E_i$ equal the average number of times a bus stops in that block group. Then the Poisson regression can be written

\begin{align}
Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
\eta_i = log(\lambda_i) &= \alpha + \sum_{n=1}^{4}x_n\beta
\end{align}

where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$.

@@

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA)
library(rstan)
library(shinystan)
library(data.table)
library(ggplot2)
library(rgdal)
library(sf)
library(tigris)
library(bayesplot)

options(mc.cores = parallel::detectCores())
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

mod_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
setDT(mod_dat)

counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
bgs <- block_groups("MN", counties, 2016)
```

@@

# The Simplest Model

Let $Y_i$ equal the average number of weekday boardings in block group $i$ in 2017, and let $E_i$ equal the average number of times a bus stops in that block group. Then the Poisson regression can be written

\begin{align}
Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
\eta_i = log(\lambda_i) &= \mu + \sum_{n=1}^{4}x_n\beta\\
\theta & \sim N(0, 1)
\end{align}

where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$ and \theta models extra-Poisson variance.

@@

```{r}
mod_dat <- mod_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(daily_stops)]

y <- mod_dat$daily_boards
E <- mod_dat$daily_stops

N <- nrow(mod_dat)

K <- 4
x <- matrix(ncol = 4, c(mod_dat$pop_density, mod_dat$emp_density, mod_dat$perc_no_veh, mod_dat$walkability))

standat1 <- list(y = y, E = E, x = x, K = K, N = N)

poisson <- "/Users/raven/Documents/honors/honors-project/stan/poisson.stan"

#poisson_fitsimp <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")

print(poisson_fitsimp, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]"), probs=c(0.025, 0.5, 0.975))
#launch_shinystan(poisson_fitsimp)
```

## Predictions
```{r}
posterior <- extract(poisson_fitsimp)
yrep <- posterior$y_rep
yrep <- yrep[sample(nrow(yrep), 50), ]

pp_check(y, yrep, ppc_dens_overlay) + 
  theme_minimal() +
  xlim(0, 200) +
  labs(title = "Actual versus fitted")
```

```{r}
mod_dat[, predicted_boards := rpois(lambda = daily_stops * exp(-1.13 + 0.06*pop_density + 0.04*emp_density + 0.16*perc_no_veh), n = nrow(mod_dat))]

mod_dat_sf <- dplyr::left_join(bgs, mod_dat)

ggplot(mod_dat_sf, aes(x=daily_boards, y = predicted_boards)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Actual versus predicted", subtitle = "Clearly overstimating outliers")

ggplot(mod_dat_sf %>% dplyr::filter(daily_boards < 500), aes(x=daily_boards, y = predicted_boards)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Actual versus predicted", subtitle = "Clearly overstimating outliers") +
  xlim(0, 500) + ylim(0, 500)
```

```{r}
a <- ggplot(mod_dat_sf %>% dplyr::filter(!is.na(daily_boards) & daily_boards != 0), aes(fill=log(daily_boards))) + 
  geom_sf(lwd = 0) +
  scale_fill_viridis_c() + 
  theme_minimal() +
  labs(title = "Observed log daily boardings") +
  theme(legend.position = 'none')

b <- ggplot(mod_dat_sf %>% dplyr::filter(!is.na(daily_boards)), aes(fill=log(predicted_boards))) + 
  geom_sf(lwd = 0) +
  scale_fill_viridis_c() + 
  theme_minimal() +
  labs(title = "Predicted log daily boardings") +
  theme(legend.position = 'none')

gridExtra::grid.arrange(a, b, ncol = 2)
```

It looks like the simple Poisson regression is beginning to capture the correct spatial pattern, though it's underestimating some areas on the suburbs.

@@

```{r}
ggplot(mod_dat_sf %>% dplyr::filter(!is.na(daily_boards)), aes(fill=log(daily_boards) - log(predicted_boards))) + 
  geom_sf(lwd = 0) +
  scale_fill_viridis_c() + 
  theme_minimal() +
  labs(title = "Residuals") +
  theme(legend.position = 'none')
```

In general, it seems the model is failing in a few ways: overestimating ridership in high-ridership block groups, and underestimating the number of block groups with low ridership.

Let's see if these residuals are spatially autocorrelated.

```{r}
library(spdep)

resid_dat <- mod_dat_sf %>% 
  dplyr::filter(!is.na(daily_boards))

nb <- poly2nb(resid_dat)
W <- nb2mat(nb, style = "B", zero.policy = TRUE)
listW <- nb2listw(nb, zero.policy = TRUE)

resid_dat <- resid_dat %>%
  dplyr::mutate(resid = daily_boards - predicted_boards)

moran.mc(resid_dat$resid, listW, nsim = 9999, zero.policy = TRUE)
```

# An Improvement

One way we could improve this model is by explicitly modeling the extra-Poisson variance (or overdispersion) which may be contributing the the poor fit above. We can add a $\theta$ parameter to address this variation as follows

\begin{align}
Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
\eta_i = log(\lambda_i) &= \mu + \sum_{n=1}^{4}x_n\beta + \theta \\
\theta & \sim N(0, 1)
\end{align}

## Results

```{r}
poisson_theta <- "/Users/raven/Documents/honors/honors-project/stan/poisson_theta.stan"

poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 7000, iter = 8000)
saveRDS(poisson_theta_fit, "/Users/raven/Documents/honors/honors-project/stan/poisson_theta_test.stan")

beepr::beep("fanfare")

print(poisson_theta_fit, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "theta[5]", "eta[5]", "lambda[5]"), probs=c(0.025, 0.5, 0.975))
launch_shinystan(poisson_theta_fit)
```

```{r}
fit <- rstan::extract(poisson_theta_fit)
test <- rpois(fit$lambda, n = 1513)

ggplot(mod_dat, aes(x=test)) +
  geom_density(color = "red") +
  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 500)
```

So this isn't great but I think it might just need more iterations? This is on the "run overnight" list.

# Testing total annual rides

```{r}
standat1 <- list(y = y*365, E = E, x = x, K = K, N = N)

poisson <- "/Users/raven/Documents/honors/honors-project/stan/poisson.stan"

poisson_fit_year <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
#poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
beepr::beep("fanfare")
print(poisson_fit_year, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]"), probs=c(0.025, 0.5, 0.975))
```

```{r}
fityear <- rstan::extract(poisson_fit_year)
testyear <- rpois(fityear$lambda, n = 1513)

ggplot(mod_dat, aes(x=testyear)) +
  geom_density(color = "red") +
  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 3000)
```

# Testing negative

```{r}
negbinom <- "/Users/raven/Documents/honors/honors-project/stan/negbinom.stan"

nb_fit <- stan(negbinom, data = standat1, warmup = 1000, iter = 2000, control = list(adapt_delta = 0.99))

print(nb_fit, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "eta[5]", "lambda[5]", "scale"), probs=c(0.025, 0.5, 0.975))

launch_shinystan(nb_fit)
```

### 2019-12-10 rmds/spatial_mod.Rmd
work from weekend at home

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA)
library(rstan)
library(shinystan)
library(data.table)
library(ggplot2)
library(rgdal)
library(sf)

options(mc.cores = parallel::detectCores())
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

mod_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
setDT(mod_dat)

counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
bgs <- block_groups("MN", counties)

spat_dat <- dplyr::left_join(bgs, mod_dat)
spat_dat <- spat_dat %>%
  dplyr::filter(!is.na(daily_boards) & !is.na(walkability) & !is.na(perc_no_veh) & !is.na(pop_density) & !is.na(emp_density))
```

@@

We can improve the Poisson regression by modeling the residuals. As we saw, there is some spatial autocorrelation in the residuals. Additionally, the distribution of $Y_i$ is overdispersed, meaning the variance is greater than we would expect from a Poisson distribution. Therefore, we'll add *two* parameters to account for both pure, uncorrelated overdispersion ($\theta$) and spatially autocorrelated variance ($\phi$).

## The Model
For areal unit $i$, the spatial model can be defined

Let $Y_i$ represent the average number of daily boardings in census block group $i$ (rounded to an integer). Then the spatial model can be defined

\begin{align}
Y_i|\theta_i & \sim \text{Poisson}(E_i\lambda_i) \\
\eta_i = \text{log}(\lambda_i) &= \mu + x\beta + ((\sqrt{\rho / s})\phi + (\sqrt{1-\rho})\theta)\sigma  \\
\phi &\sim N(0, Q^{-1}) \\
\theta &\sim N(0, n) \\
\rho & \sim \text{beta}(1/2, 1/2) \\
\sigma &\sim \text{exponential }

\end{align}

where

* E is an offset term, sometimes also called the baseline or "expected value" of Y (same as Poisson regression)
* $\lambda_i$ is the relative "risk", and $\eta_i$ is the log relative risk
* $\mu$ is the overall "risk" level, an intercept term
* $x\beta$ are covariates which vary spatially
* $\phi$ is the intrinsic conditional autoregressive (ICAR) model which accounts for spatially structured errors. In this case, we are using the BYM2 model
* $\theta$ accounts for pure overdispersion. This parameter is the complement to $\phi$ and accounts for non-spatial errors. 
* $\rho$ is a mixing parameter. It determines how much variance comes from spatially correlated versus independent error terms. The introduction of $\rho$ is a key difference between the BYM and BYM2 models. 
* $s$ is a scaling factor such that $Var(\phi_i) \approx 1$. We set this variance so that $\sigma$ is the actual standard deviation of the errors.
* $n$ is the number of fully connected subgraphs in the neighborhood structure. $n=1$ for study areas with no regions with 0 neighbors. 
* $\sigma \geq 0$ is the overall standard deviation for combined error terms

# Results

## One covariate, Ramsey County
```{r}
spat_dat <- spat_dat %>% 
  dplyr::filter(COUNTYFP == 123)

y = spat_dat$daily_boards
E = spat_dat$daily_stops

K = 1
x = matrix(ncol = 1, c(spat_dat$emp_density))

nb <- poly2nb(spat_dat)

source("nb_helper_funs.R")
nbs <- nb2graph(nb)
N <- nbs$N
node1 <- nbs$node1
node2 <- nbs$node2
N_edges <- nbs$N_edges
scaling_factor <- scale_nb_components(nb)[1]

standat1 <- list(N = N, N_edges = N_edges, node1 = node1, node2 = node2, y = y, E = E, scaling_factor = scaling_factor, K = K, x = x)

bym2 <- "/Users/raven/Documents/honors/honors-project/stan/bym2.stan"

#bym2_fit1 <- stan(bym2, data = standat1, warmup = 5000, iter = 6000, verbose = T)
#saveRDS(bym2_fit1, "/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")

bym2_fit1 <- readRDS("/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
```

```{r}
print(bym2_fit1, pars=c("beta0", "betas[1]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975))
#launch_shinystan(bym2_fit1)
```

I'm not sure if the problem with this is that I didn't run it enough, it's parameterized poorly, etc. Actually only took ~10 minutes to run.

# All four covariates, Ramsey

Using not enough iterations right now so I can see if it works
```{r}
x2 <- matrix(ncol = 4, c(spat_dat$pop_density, spat_dat$emp_density, spat_dat$perc_no_veh, spat_dat$walkability))
K2 <- 4

standat2 <- list(N = N, N_edges = N_edges, node1 = node1, node2 = node2, y = y, E = E, scaling_factor = scaling_factor, K = K2, x = x2)

#bym2_fit2 <- stan(bym2, data = standat2, warmup = 2000, iter = 3000, thin = 1, verbose = T)

#saveRDS(bym2_fit2, "/Users/raven/Documents/honors/honors-project/fits/bym2_fit2.RDS")

bym2_fit2 <- readRDS("/Users/raven/Documents/honors/honors-project/fits/bym2_fit2.RDS")
```

```{r}
#launch_shinystan(bym2_fit2)
print(bym2_fit2, pars=c("beta0", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975))
```

Everything except $\rho$ seems to be alright..
